### 2016-12-07 paper58/paper58.Rmd
new updates - Climatrans
updated paper58.Rmd

*Benestad et al - Upper-limit estimation of future precipitation return-values* 
-----------------------------------
title: "Analysis and calculations in manuscript"
author: "Rasmus Benestad"
date: "April 4, 2016"
output: pdf_document
---

@@

## Start the esd-library:
library(esd)
## Information about the system and session
Sys.info()
sessionInfo()
```

@@

**Simple functions which estimate various statistics**

  Functions to extract the numbers for mean,min,max,wettest month,driest month, estimate percentiles assuming an exponential distribution, skill-scores, and regression coefficients. These are used to make the R-code simpler, shorter, and improve clarity.

@@

**Generation of graphics/maps**

@@

@@

@@

Now the data is ready for the analysis. Calibrate the regression models and extract the regression coeffieients. Use 'apply' to speed up the process for multiple stations.

@@

Need to collect aggregated results based on the ensembles of GCM projections.

@@

## derive time series for each location:
Z.rcp4.5 <- lapply(Vx,'muproject',rcp4.5)
Z.rcp8.5 <- lapply(Vx,'muproject',rcp8.5)
Z.rcp2.6 <- lapply(Vx,'muproject',rcp2.6)
t <- index(rcp4.5)
```
Extract the statistics of the different emission scenarios and time slices

@@

figlab('Figure 3')
```

**Supporting material/analysis (SM)**

Extract the essential information

@@

##** ----------- Supporting Material ------------------------**

The supporting material includes a number of diagnostics and plots that chencks a number of assumptions made in this study.

##**figure SM2**

Presents the statistics of $R^2$ from the model calibration based on the different locations.

@@

## Plot the statistics of R2:
hist(100*as.numeric(r2),breaks=seq(0,100,by=5),lwd=2,col=rgb(0,0.3,0.5),
     xlab=expression(paste(R^2,' (%)')),freq=TRUE,
     main="Summary of regression scores")
grid()
figlab('Figure SM2')
```
##**Figure SM12**

@@

Historical trends in the wet-day mean preciptation

```{r}
par(bty='n',col.sub='grey',cex.sub=0.8)
plot(trend.mux,trend.pre,pch=19,col=rgb(0.6,0.2,0,0.3),cex=1.5,
     main=expression(paste('Trends in ',mu,': observed and predicted upper limit')),
     xlab='Observed trend (mm/decade)',ylab='predicted trend (mm/decade)',
     xlim=xylim,ylim=xylim,
     sub=paste('Mean correlation for local year-to-year variations over t=[',
       start(MUx),',',end(MUx),
       '] is ',round(mean(r.eval),2),' (',round(quantile(r.eval,0.05),2),', ',
       round(quantile(r.eval,0.95),2),
       ')',sep='') )
grid()

```{r}
## Map showing trends in mu
map(MU,FUN='trend',colbar=list(breaks=seq(-1,1,length=21),rev=TRUE),cex=cex)
figlab('Figure SM4',ypos=0.999)
figlab(expression(paste('Trend in ',mu,' (mm/day per decade)')),xpos=0.5,ypos=0.999)
```
##**Figure SM5**

```{r}
## 99-percentile wet-day mean and typical wet-day frequency
## The Tellus paper on specification approximately annual maximum.
## Prob(X>x) for annual maximum for 24hr precip
## Pr(X > x) = 1/365.25 = fw*(1-p): p = 1 - 1/(365.25*fw)
par(bty='n')
W <- apply(fw,2,FUN='muclim')
hist(100*(1-1/(W[1,]*365.25)),col='steelblue',lwd=2,breaks=seq(90,100,by=0.1),
     main='Wet-day percentile for annual maximum 24-precipitation',
     xlab='p (%)')
grid()
figlab('Figure SM5')
```

```{r}
## Statistics of trend in wet-day frequency
print('Wet-day frequency statistics')
hist(fw.trend,breaks=seq(-50,50,by=1),col='grey',lwd=2,
     main='Trend in wet-day frequency',
     xlab=expression(paste(f[w],' (%/decade)')))
grid()
figlab('Figure SM6')
```
##**Figure SM7**

@@

Additional analysis testing the connection between local temperature and wet-day mean precipitation

```{r}
## Regression analysis between the wet-day mean mu and the mean temperature.

corhalf <- function(x) {
  n <- length(x)
  x1 <- x[1:(n/2)]; x2 <- x[(n/2+1):n]
  ok <- is.finite(x1) & is.finite(x2)
  r <- cor(x1[ok],x2[ok])
  return(r)
}
```

```{r}
## CLARIS
load('claris.Tx.rda')
load('claris.Pr.rda')

@@

@@

```{r}
## Aggregate annual statistics based on the combined data sources:
MU <- combine(mu,annual(Pr1,FUN='wetmean'),annual(Pr2,FUN='wetmean'))
FW <- combine(fw,annual(Pr1,FUN='wetfreq'),annual(Pr2,FUN='wetfreq'))
T2M <- combine(t2m,annual(Tx1,FUN='mean'),annual(Tx2,FUN='mean'))
nval <- apply(coredata(MU),2,'nv')
attr(T2M,'variable') <- 't2m'
es <- C.C.eq(T2M)

#map(MU,FUN='mean',cex=0)
#map(MU,FUN='trend',cex=0,colbar=list(rev=TRUE,breaks=seq(-1,1,length=21)))
#map(FW,FUN='trend',cex=0,colbar=list(rev=TRUE,breaks=seq(-0.01,0.01,length=21)))
calmu <- data.frame(x=as.numeric(apply(coredata(es),2,'mean',na.rm=TRUE)),
                    y=as.numeric(apply(coredata(MU),2,'mean',na.rm=TRUE)),
                    fw=as.numeric(apply(coredata(FW),2,'mean',na.rm=TRUE)),
                    z=alt(MU),lat=lat(MU),lon=lon(MU),nval=nval)
premu <- calmu; premu$x[0] <- 0
model.mutx <- lm(y ~ x, weights=fw,data=calmu)
print(summary(model.mutx))
data(geoborders)
Fw <- apply(coredata(fw),2,'sum',na.rm=TRUE)
```

@@

```{r}
## Estimate the year-by-year correlation in es and mu

>>>>

*Benestad et al - Simple and approximate estimation of future precipitation return-values* 
-----------------------------------
title: "Analysis and calculations in manuscript"
author: "Rasmus Benestad"
date: "April 4, 2016"
output: pdf_document
---

@@

## Start the esd-library:
library(esd)
## Information about the session
sessionInfo()
```

@@

**Simple functions to estimate various statistics**

  Functions to extract the numbers for mean, min, max, wettest month, driest month, estimate percentiles assuming an exponential distribution, skill-scores, and regression coefficients. These are used to make the R-code simpler, shorter, and improve clarity.

@@

**Generation of graphics**

@@

##**Calculations and analysis**

@@

**Access and prepare data**

@@

**Statistical downscaling and analysis**

Now the data is ready for the analysis. Calibrate the regression models and extract the regression coefficients. Use 'apply' to speed up the process for multiple stations.

@@

Collect aggregated GCM projections (CMIP5) and apply the regression models to the ensembles.

@@

## derive time series for each location:
Z.rcp4.5 <- lapply(Vx,'muproject',rcp4.5)
Z.rcp8.5 <- lapply(Vx,'muproject',rcp8.5)
Z.rcp2.6 <- lapply(Vx,'muproject',rcp2.6)
t <- index(rcp4.5)
```
Extract statistics of the wet-daymean associated with different emission scenarios and time slices and calculate the changes from the present (2010) to the far future (2100).

@@

figlab('Figure 3')
```
##** ----------- Supporting Material ------------------------**

The supporting material includes diagnostics and plots that checks a number of assumptions made in this study.

**Reorganise data and statistical information**
Extract and reorganise essential information from previous analysis that will be used in the SM.

@@

Figure SM1 shows a test of the assumption that that the wet-day daily precipitation is approximately exponentially distributed, by comparing the actual percentiles with quantiles estimated for different samples 
with different annual mean wet-day precipitation using the formula for exponentially distributed data: $q_p = -ln(1-p)\mu$.

##**Figure SM2**

Figure SM2 presents the $R^2$ statistics from the model calibration (i.e., the regression between the seasonal cycles of $\mu$ and $e_s$) for different locations.

@@

## Plot the statistics of R2:
hist(100*as.numeric(r2),breaks=seq(0,100,by=5),lwd=2,col=rgb(0,0.3,0.5),
     xlab=expression(paste(R^2,' (%)')),freq=TRUE,
     main="Summary of regression scores")
grid()
figlab('Figure SM2')
```

Figure SM3 compares the observed and predicted historical trends in the wet-day mean preciptation.

```{r}
par(bty='n',col.sub='grey',cex.sub=0.8)
plot(trend.mux,trend.pre,pch=19,col=rgb(0.6,0.2,0,0.3),cex=1.5,
     xlab=expression(paste('Observed trend in ',mu,' (mm/decade)')),
     ylab=expression(paste('Predicted trend in ',mu,'(mm/decade)')),
     xlim=xylim,ylim=xylim,
     sub=paste('Mean correlation for local year-to-year variations over t=[',
       start(MUx),',',end(MUx),
       '] is ',round(mean(r.eval),2),' (',round(quantile(r.eval,0.05),2),', ',
       round(quantile(r.eval,0.95),2),
       ')',sep='') )
grid()

Figure SM4 shows the historical trends in the wet-day mean preciptation.

```{r}
## Map showing trends in mu
map(MU,FUN='trend',colbar=list(pal="burd",breaks=seq(-1,1,length=21)),cex=cex)
figlab('Figure SM4',ypos=0.999)
figlab(expression(paste('Trend in ',mu,' (mm/day per decade)')),xpos=0.5,ypos=0.999)
```

## Figure SM5 is not used in either the main paper of supplementary material
<##**Figure SM5**>

#Figure SM5 is a histogram of the  shows the historical trends in the wet-day mean preciptation.

#```{r}
## 99-percentile wet-day mean and typical wet-day frequency
## Prob(X>x) for annual maximum for 24hr precip
## Pr(X > x) = 1/365.25 = fw*(1-p): p = 1 - 1/(365.25*fw)
#par(bty='n')
#W <- apply(fw,2,FUN='muclim')
#hist(100*(1-1/(W[1,]*365.25)),col='steelblue',lwd=2,breaks=seq(90,100,by=0.1),
#     main='Wet-day percentile for annual maximum 24-hr precipitation',
#     xlab='p (%)')
#grid()
#figlab('Figure SM5')
#```

Figure SM6 displays a histogram of historical trends in the wet-day frequency for the period... for the stations in Europe shown in Figure SM7.

```{r}
## Statistics of trend in wet-day frequency
print('Wet-day frequency statistics')
hist(fw.trend,breaks=seq(-50,50,by=1),col='grey',lwd=2,
     main='Trend in wet-day frequency',
     xlab=expression(paste(f[w],' (%/decade)')))
grid()
figlab('Figure SM6')
```

##**Figure SM7**

Figure SM7 shows the spatial distribution of the historical trends in wet-day frequency on a map.

##**Figure SM12**

@@

@@

Additional analysis testing the connection between local temperature and wet-day mean precipitation.

**Regression analysis between the wet-day mean mu and the mean temperature**

Load South American temperature and precipitation data from the climate database CLARIS and COST-VALUE(?). 
```{r}
## CLARIS
load('claris.Tx.rda')
load('claris.Pr.rda')

@@

Load and prepare temperature and precipitation data from North America from the GDCN database.

@@

Combine and aggregate the observational data from North America and South America, and then fit a regression model based on comparing the annual mean $\mu$ and $e_s$ at the different stations.  
```{r}
## Aggregate annual statistics based on the combined data sources:
MU <- combine(mu,annual(Pr1,FUN='wetmean'),annual(Pr2,FUN='wetmean'))
FW <- combine(fw,annual(Pr1,FUN='wetfreq'),annual(Pr2,FUN='wetfreq'))
T2M <- combine(t2m,annual(Tx1,FUN='mean'),annual(Tx2,FUN='mean'))
nval <- apply(coredata(MU),2,'nv')
attr(T2M,'variable') <- 't2m'
es <- C.C.eq(T2M)

calmu <- data.frame(x=as.numeric(apply(coredata(es),2,'mean',na.rm=TRUE)),
                    y=as.numeric(apply(coredata(MU),2,'mean',na.rm=TRUE)),
                    fw=as.numeric(apply(coredata(FW),2,'mean',na.rm=TRUE)),
                    z=alt(MU),lat=lat(MU),lon=lon(MU),nval=nval)
premu <- calmu; premu$x[0] <- 0
model.mutx <- lm(y ~ x, weights=fw,data=calmu)
print(summary(model.mutx))
data(geoborders)
Fw <- apply(coredata(fw),2,'sum',na.rm=TRUE)
```

@@

```{r}
## Estimate the year-by-year correlation in es and mu
corhalf <- function(x) {
  n <- length(x)
  x1 <- x[1:(n/2)]; x2 <- x[(n/2+1):n]
  ok <- is.finite(x1) & is.finite(x2)
  r <- cor(x1[ok],x2[ok])
  return(r)
}
